---
title: "Get RT data from chronolex"
output: pdf_document
---

Based on `process.words.R` available from the French lexicon project results folder:
https://sites.google.com/site/frenchlexicon/results
download >> "French Lexicon Project trial-level results with R scripts.zip"

```{r message=FALSE}
rm(list=ls())
# change directory to chronolex folder
setwd("./chronolex")
a<-read.table("results.utf8.txt", as.is=T, encoding="utf-8", sep='\t',
              col.names=c('sujet','num','mot','rt','acc','typemot','ordre','ordresuite'))
a <- a[order(a$sujet),]
a$sujet <- as.factor(a$sujet)
#a$sujet = paste("subj",a$sujet,sep="")

# number of subjects
length(unique(a$sujet))
```

```{r eval=false}
# suppress the pseudowords
b = subset(a, typemot=="mot")
(n = nrow(b))
```


```{r}
# plot accuracy ~ rt per subject 
rtsuj = tapply(a$rt, a$sujet, median)
accsuj = tapply(a$acc, a$sujet, mean)
ntrials = tapply(a$rt, a$sujet, length) # one P with 500 trials per condition

plot(rtsuj, accsuj, main="RT & accuracy (one point per subject)") 
rtcutoff <- 1500 #1100
acccutoff <- .4 #.75
abline(h=acccutoff)
abline(v=rtcutoff)
```
Check participants with RT > 1500
```{r}
list.p <- unique(a$sujet)
pp <- list.p[rtsuj>1500]
hist(a$rt[a$sujet==pp[1]],breaks=50,xlim=c(0,2500))
```

```{r}
list.p <- unique(a$sujet)
pp <- list.p[rtsuj>1500]
hist(a$rt[a$sujet==pp[2]],breaks=50,xlim=c(0,2500))
```

```{r}
list.p <- unique(a$sujet)
pp <- list.p[rtsuj>1500]
hist(a$rt[a$sujet==pp[3]],breaks=50,xlim=c(0,2500))
```

```{r}
list.p <- unique(a$sujet)
pp <- list.p[rtsuj>1500]
hist(a$rt[a$sujet==pp[4]],breaks=50,xlim=c(0,2500))
```

Clearly some participants with low accuracy and high RT did not follow the task instructions.

```{r}
# plot accuracy ~ rt 3rd quartile per subject 
qrtsuj = tapply(a$rt, a$sujet, quantile, probs = 0.75)
accsuj = tapply(a$acc, a$sujet, mean)
ntrials = tapply(a$rt, a$sujet, length) # one P with 500 trials per condition

plot(qrtsuj, accsuj, main="RT & accuracy (one point per subject)") 
qrtcutoff <- 1500 #1100
acccutoff <- .4 #.75
abline(h=acccutoff)
abline(v=rtcutoff)
```

Also, several participants have very unusual RT distributions, with a second bump of late RT.

```{r}
list.p <- unique(a$sujet)
pp <- list.p[qrtsuj>1500]
hist(a$rt[a$sujet==pp[1]],breaks=50,xlim=c(0,2500))
# hist(a$rt[a$sujet==631],breaks=50,xlim=c(0,2500))
```
```{r}
list.p <- unique(a$sujet)
pp <- list.p[qrtsuj>1500]
hist(a$rt[a$sujet==pp[2]],breaks=50,xlim=c(0,2500))
# hist(a$rt[a$sujet==876],breaks=50,xlim=c(0,2500))
```

```{r eval=TRUE}
# select subjects according to cutoffs
goodsubjects <- (rtsuj<rtcutoff) & (accsuj>acccutoff) & (ntrials>1500) & (qrtsuj<qrtcutoff)
a <- subset(a, goodsubjects[a$sujet]==TRUE)
length(unique(a$sujet))
nrow(a)
```

```{r}
# number of trials per item (including errors)
ntrials = tapply(a$rt, a$mot, length)
```

```{r}
# number of trials per participant (including errors)
aw = subset(a, typemot=="mot")
ntrials = tapply(aw$rt, aw$sujet, length)
summary(ntrials)
anw = subset(a, typemot=="nonmot")
ntrials = tapply(anw$rt, anw$sujet, length)
summary(ntrials)

ntrials = tapply(a$rt, list(a$sujet,a$typemot), length)
summary(ntrials)
```

```{r}
summary(tapply(aw$rt, aw$sujet, median))
summary(tapply(anw$rt, anw$sujet, median))

par(mfcol=c(1,2))
hist(tapply(aw$rt, aw$sujet, median),breaks=50,xlim=c(0,2000),main="Median RT by part.: words")
hist(tapply(anw$rt, anw$sujet, median),breaks=50,xlim=c(0,2000),main="Median RT by part.: non-words")
```


```{r eval=FALSE}
# suppress trials with wrong responses
n = nrow(b)
b = subset(b, b$acc==1)
1-(nrow(b)/n) # proportion of rejected datapoints

nhits = tapply(b$rt, b$mot, length)
```

```{r}
library(tibble)
# save data frame
b <- subset(a, a$typemot=="mot" | a$typemot=="nonmot") 
b$typemot <- factor(b$typemot)
library(plyr)
b$typemot <- revalue(b$typemot, c("mot"="word", "nonmot"="non-word"))
flp <- tibble(`participant` = as.factor(b$sujet),
              `rt` = b$rt,
              `acc` = b$acc,
              `condition` = b$typemot)
# get rid of missing participants
flp$participant <- factor(flp$participant)
save(flp, file="french_lexicon_project_rt_data.RData")
```

```{r}
graphics.off()
proc.time()
```



